Gusto: a toolkit for compatible finite element dynamical cores

### 2024-07-24 source/paper.tex
some progress on FML section

\section{Governing Equations}
\label{sec: governing}
The aim of this section is to introduce three governing equation sets
commonly used in the development of weather and climate models, along
with enough detail about the finite element discretiations provided by
Gusto to demonstrate the similarities between them. These similarities
guide the design of the code structure described in the next section
and motivate the use of Form Manipulation Language (FML) that
underlies the flexibility of Gusto, a detailed example of which will
be given in section \ref{sec: FML}. The convention in Gusto is to
write the equations in residual form, meaning that all the terms
appear on the left hand side. We will follow this convention below in
contrast to many papers which present the equations with the pressure
gradient term and any forcing terms on the right hand side.

@@

The shallow water equations describe the flow of a shallow layer of
fluid subject to gravitational and, optionally, rotational forces. The
prognostic variables are the two horizontal velocity components
$\MM{u} = (u, v)$ and the fluid depth $D = H + h - b$ where $H$ is the
undisturbed depth of the fluid layer, $h$ is the perturbation to the
fluid depth and $b$ is the bottom topography, if present. The
equations are:

\begin{align}
  \pp{\MM{u}}{t} + \MM{u}\cdot\nabla\MM{u} + f\hat{\MM{k}}\times\MM{u} + g\nabla (D+b) &= 0, \\
  \pp{D}{t} + \nabla\cdot(D\MM{u}) &= 0,
\end{align}
where $f$ is the Coriolis parameter, $\hat{\MM{k}}$ is the unit vector
pointing upwards and $g$ is the acceleration due to gravity. The
nonlinear velocity advection term can be replaced by the sum of a
vorticity-based term and the gradient of the kinetic energy. This is
known as the vector invariant form and it enables the construction of
schemes that conservation energy and/or enstrophy
\citep{mcrae2014energy, bauer2018energy, wimmer2020energy,
  wimmer2021energy}. However, this form can exhibit numerical
instabilities \citep{bell2017numerical} and it is not yet clear which
form is preferable so in Gusto we offer both forms of the equations.

\subsection{Compressible Boussinesq Equations}
The compressible Boussinesq equations are commonly used in ocean
modelling. They are useful from a numerical discretisation perspective
because they incorporate compressibility effects without the
complexity of discretising the nonlinear pressure gradient terms that
we shall see in the compressible Euler equations. We write these
equations in terms of the prognostic variables $\MM{u}$, $b$ and
$p$ where $\MM{u}=(u, v, w)$ is now a three dimensional velocity
field, $b$ is the fluid buoyancy and $p$ is pressure. In terms of
these variables, the equations are:

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  2\MM{\Omega}\times \MM{u} + \nabla p + b\hat{\MM{k}} & = 0, \\
  \pp{p}{t} + c^2\nabla\cdot\MM{u} & = 0, \\
  \pp{b}{t} + \MM{u}\cdot\nabla b & = 0,
\end{align}
where $\MM{\Omega}$ is the rotation vector of the domain,
$\hat{\MM{k}}$ is the unit vector pointing upwards and $c$ is the
acoustic wave speed.

\subsection{Compressible Euler Equations}
The dynamical core at the heart of any weather or climate model solves
the rotating compressible Euler equations that model atmospheric
flow. We write these equations in terms of the prognostic variables
$\MM{u}$, $\rho$ and $\theta$ where $\MM{u}=(u, v, w)$ is now a three
dimensional velocity field, $\rho$ is the fluid density and $\theta$
is the potential temperature. The pressure gradient term is written in
terms of the Exner pressure $\Pi$ which is calculated through an
equation of state. In terms of these variables, the equations are:

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  2\MM{\Omega}\times \MM{u} + c_p\theta\nabla \Pi + g\hat{\MM{k}} & = 0, \\
  \pp{\rho}{t} + \nabla\cdot(\MM{u}\rho) & = 0, \\
  \pp{\theta}{t} + \MM{u}\cdot\nabla\theta & = 0, \\
  \Pi^{(1-\kappa)/\kappa} & = \frac{R}{p_0}\rho\theta, & 
\end{align}
where $\Pi$ is the Exner pressure, $p_0$ is a reference pressure,
typically taken to be *, $R$ is ** and $\kappa$ is **.

@@

\subsection{Pressure gradient term}
The pressure gradient term is discretised in a similar way for each
equation set, with some additional considerations required for the
nonlinear form of this term in the compressible Euler equations. Since
the pressure (or depth, for shallow water) field is discretised in
$\mathbb{V}_p \subset L_2$, the gradient is not defined globally and
the term must be integrated by parts. In the linear case, using the
notation for the compressible Boussinesq equations, the weak form is

@@

\begin{equation}
  c_p\int_\Omega\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_\Omega\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\Gamma_v}\jump{\theta\MM{w}\cdot\MM{n}}\avg{\Pi} \diff S,
\end{equation}
where $\Gamma_v$ denotes the vertical element facets, across which
$\theta$ is discontinuous. Here we have chosen, as in \ref{}, to take
the average value of $\Pi$ across the facets, $\avg{\Pi} =
(\Pi^++\Pi^-)/2$, where the $+$ and $-$ indicate each side fo the
facet. The brackets $\jump{q} = q^+-q^-$ denote the jump in a value
across the facet and the integral over the horizontal facets vanishes
since both $\theta$ and $\MM{w}\cdot\MM{n}$ are continuous across
these facets so the jump is zero.

@@

** Don't do whole equation set - go term by term and try to write as
few equations as possible: transport, pressure gradient, diffusion

@@

\JScomment{Text here depends on what the schematic looks like - this
  paragraph should refer to the schematic and explain all aspects} The
Gusto model comprises a domain and a prognostic equation set. The
domain holds the mesh and the finite element function spaces along
with any other information about the space-time domain such as the
coordinates and the timestep. The Prognostic equation sets are each
defined in a python class, making use of the common finite element
forms for the transport and difffusion terms. The timeloop class
contains the definition of the timestepping procedure and the
\texttt{run} method that loops over the timesteps. It also sets up
diagnostic fields and calls the IO as required. The simplest timeloop
class will just apply one time discretisation method to the equation
set but it is common in weather and climate prediction models to split
the different types of terms and apply a time discretisation
separately to each type of term. For example, the transport and
forcing terms might be solved for separately, such as in **, which
enables explicit transport schemes to be applied to each field. **
something about SIQN structure. Terms related to the physics are
solved for in the appropriate part of the iteration loop, with physics
processes that occur on `fast' timescales evaluated more frequently
and `slower' physics processes evaluated just once at the end of the
timestep. The choice of where to evaluate the physics processes in the
iteration is also based on the computational cost and the best way to
do this is still an open research question (**ref).

@@

\section{Form Manipulation Language}
\label{sec: FML}
The discussion of the time discretisation so far has been general i.e.
not specific to any particular equation set. This is also true of the
implementation of the timestepping schemes. This enables flexibility
and reproducibility (**elaborate). This functionality is enabled
through the form manipulation language (FML) which is used to
associate a label to each term in the residual form defined in the
equation class. The timestepping methods are then able to reason about
which manipulations to apply to which forms by applying filters. A
simple example would be a splitting method that evaluates transport
terms and diffusion terms separately. The full residual is passed to
the transport scheme which then applies a filter to drop all terms
that are not labelled as transport; similarly for diffusion. We
illustrate this procedure in figure \ref{fig:advdiff} where we show
the definition of the residual for the advection diffusion equation
for a scalar variable $q$ with a prescribed transport velocity
$\MM{c}$ and diffusivity $\kappa$

\begin{equation}
  q_t + \MM{c}\cdot\nabla q - \kappa\nabla^2 q = 0,
\end{equation}
with the terms labelled either \texttt{time\_derivative},
\texttt{transport} or \texttt{diffusion}. The timestepping loop then
applies a step of forward Euler to the transport term and backwards
Euler to the diffusive term, i.e. the time discretisation is

\begin{align}
  q^* &= q^n - \Delta t (\MM{c}\cdot\nabla) q^n, \\
  q^{n+1} &= q^* + \Delta t \kappa \nabla^2 q^{n+1}.
\end{align}

@@

To include:
\begin{itemize}
\item terminator toy (Tim)
\item shallow water on different grids and with different HDiv
  families (Nell) e.g. Williamson 5 cubed sphere PV and Galewsky
  icosahedral sphere PV, high res (ref 6)
\item convergence with different RK schemes / multistage schemes using
  Williamson 2 or 6, L2 and Linf errors for D and u (2 plots, one L2,
  one Linf) (Alex)
\item vertical slice Euler - Straka, different orders, illustrates
  recovery for diffusion (Tom)
\item 3D moist bubble with rain (Tom/Alex)
\item spherical gravity wave, Boussinesq (Jemma)
\item spherical baroclinic wave, Euler - whatever people usually plot
  (horizontal slice pressure contours, temperature shadng) (Dan)
\item Held Suarez??
\end{itemize}

@@

\subsection{Shallow water}
The standard suite of shallow water test cases originally proposed in
\citet{williamson1992standard} provides a common starting point for
testing new discretisations. This test suite is usually augmented by
the unstable barotropic jet described in \citet{galewsky2004initial}

>>>>

\section{Governing Equations}
\label{sec: governing}
The aim of this section is to introduce three governing equation sets
commonly used in the development of discretisation methods for weather
and climate models, along with enough detail about the finite element
discretiations provided by Gusto to demonstrate the similarities
between them. These similarities guide the design of the code
structure described in the next section and motivate the use of Form
Manipulation Language (FML) that underlies the flexibility of Gusto, a
detailed example of which will be given in section \ref{sec: FML}. The
convention in Gusto is to write the equations in residual form,
meaning that all the terms appear on the left hand side. We will
follow this convention below in contrast to many papers which present
the equations with the pressure gradient term and any forcing terms on
the right hand side.

@@

The shallow water equations describe the flow of a shallow layer of
fluid subject to gravitational and, optionally, rotational forces. The
prognostic variables are the two horizontal velocity components
$\MM{u} = (u, v)$ and the fluid depth $D = H + h - b$ where $H$ is the
undisturbed depth of the fluid layer, $h$ is the free surface
deviation from $H$, and $b$ is the bottom topography, if present. The
equations are:

\begin{align}
  \pp{\MM{u}}{t} + (\MM{u}\cdot\nabla)\MM{u} + f\hat{\MM{k}}\times\MM{u} + g\nabla (D+b) &= 0, \\
  \pp{D}{t} + \nabla\cdot(D\MM{u}) &= 0,
\end{align}
where $f$ is the Coriolis parameter, $\hat{\MM{k}}$ is the unit vector
pointing upwards and $g$ is the acceleration due to gravity.

\subsection{Compressible Boussinesq Equations}
The compressible Boussinesq equations are commonly used in ocean
modelling. They are useful from a numerical discretisation perspective
because they incorporate compressibility effects without the
complexity of discretising the nonlinear pressure gradient terms that
we shall see in the compressible Euler equations. We write these
equations in terms of the prognostic variables $\MM{u}$, $b$ and $p$
where $\MM{u}$ is a three dimensional velocity field, $b$ is the fluid
buoyancy and $p$ is pressure. In terms of these variables, the
equations are:

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  2\MM{\Omega}\times \MM{u} + \nabla p + b\hat{\MM{k}} & = 0, \\
  \pp{p}{t} + c^2\nabla\cdot\MM{u} & = 0, \\
  \pp{b}{t} + (\MM{u}\cdot\nabla) b & = 0,
\end{align}
where $\MM{\Omega}$ is the rotation vector of the domain,
$\hat{\MM{k}}$ is the unit vector pointing upwards and $c$ is the
acoustic wave speed. The velocity $\MM{u}$ can be two or three
dimensional, depending on whether the domain is a vertical slice
described by $(x, z)$ coordinates, of a three dimensional domain
described by $(x, y, z)$ coordinates.

\subsection{Compressible Euler Equations}
The dynamical core at the heart of any weather or climate model solves
the rotating compressible Euler equations that model atmospheric
flow. We write these equations in terms of the prognostic variables
$\MM{u}$, $\rho$ and $\theta$ where $\MM{u}$ is the velocity field,
$\rho$ is the fluid density and $\theta$ is the potential
temperature. The pressure gradient term is written in terms of the
Exner pressure $\Pi$ which is calculated through an equation of
state. In terms of these variables, the equations are:

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  2\MM{\Omega}\times \MM{u} + c_p\theta\nabla \Pi + g\hat{\MM{k}} & = 0, \\
  \pp{\rho}{t} + \nabla\cdot(\MM{u}\rho) & = 0, \\
  \pp{\theta}{t} + (\MM{u}\cdot\nabla)\theta & = 0, \\
  \Pi^{(1-\kappa)/\kappa} & = \frac{R}{p_0}\rho\theta, & 
\end{align}
where $\Pi$ is the Exner pressure, $p_0$ is a reference pressure,
typically taken to be *, $R$ is ** and $\kappa$ is **. The velocity
$\MM{u}$ can be two or three dimensional, depending on whether the
domain is a vertical slice described by $(x, z)$ coordinates, of a
three dimensional domain described by $(x, y, z)$ coordinates.

In all of the equation sets described here, the nonlinear velocity
advection term $(\MM{u}\cdot\nabla)\MM{u}$ can be replaced by the sum
of a vorticity-based term and the gradient of the kinetic energy. This
is known as the vector invariant form and it enables the construction
of schemes that conservation energy and/or enstrophy
\citep{mcrae2014energy, bauer2018energy, wimmer2020energy,
  wimmer2021energy}. However, this form can exhibit numerical
instabilities \citep{bell2017numerical} and it is not yet clear which
form is preferable so in Gusto we offer both forms of the equations.

@@

\subsection{Pressure gradient term}
The pressure gradient term is discretised in a similar way for each
equation set, with some additional considerations required for the
nonlinear form of this term in the compressible Euler equations. Since
the pressure (or depth, for shallow water) field is discretised in
$\mathbb{V}_p \subset L_2$, meaning that it is discontinuous between
cells, the gradient is not defined globally and the term must be
integrated by parts. In the linear case, using the notation for the
compressible Boussinesq equations, the weak form is

@@

\begin{equation}
  c_p\int_\Omega\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_\Omega\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\Gamma_v}\jump{\theta\MM{w}\cdot\MM{n}}\avg{\Pi} \diff S,
\end{equation}
where $\Gamma_v$ denotes the vertical element facets, across which
$\theta$ is discontinuous. Here we have chosen, as in \citet{}, to take
the average value of $\Pi$ across the facets, $\avg{\Pi} =
(\Pi^++\Pi^-)/2$, where the $+$ and $-$ indicate each side of the
facet. The brackets $\jump{q} = q^+-q^-$ denote the jump in a value
across the facet and the integral over the horizontal facets vanishes
since both $\theta$ and $\MM{w}\cdot\MM{n}$ are continuous across
these facets so the jump is zero.

@@

For the simulations presented here we do not require viscous or
diffusive terms for stability, but sometimes they are required to
produce converged solutions, ie. so that refining the grid does not
lead to smaller and smaller scale motions. Here we present the symmetric
interior penalty discretisation method used for these terms in Gusto.

\begin{equation}
  \nu\int_\Omega\MM{w}\cdot\nabla^2\MM{u}\diff x = \nu\int_\Omega\nabla\MM{w}:\nabla\MM{u}\diff x
\end{equation}

@@

\JScomment{This paragraph should refer to the schematic and explain
  all aspects} The Gusto model comprises a domain and a prognostic
equation set. The domain holds the mesh and the finite element
function spaces along with any other information about the space-time
domain such as the coordinates and the timestep. The prognostic
equation sets are each defined in a python class, making use of the
common finite element forms for the transport and difffusion terms to
set up the residual (defined in section \ref{}) in weak form. At this
point, no particular discretisation of the transport or diffusion
forms is specified. This happens in the \texttt{SpatialMethod} class
which can be used to apply the methods described in section \ref{} to
the relevant terms. Any physics source terms, such as **, are added by
calling the \texttt{Physics} class that defines them. The
\texttt{TimeStepper} class contains the definition of the timestepping
procedure and the \texttt{run} method that loops over the
timesteps. It also sets up diagnostic fields and calls the IO as
required. The simplest timeloop class will just apply one time
discretisation method to the equation set but it is common in weather
and climate prediction models to split the different types of terms
and apply a time discretisation separately to each type of term. For
example, the transport and forcing terms might be solved for
separately, such as in **, which enables explicit transport schemes to
be applied to each field. ** something about SIQN structure. Terms
related to the physics are solved for in the appropriate part of the
iteration loop, with physics processes that occur on `fast' timescales
evaluated more frequently and `slower' physics processes evaluated
just once at the end of the timestep. The choice of where to evaluate
the physics processes in the iteration is also based on the
computational cost and the best way to do this is still an open
research question (**ref).

\begin{itemize}
\item what about initialisation and diagnostics?
\end{itemize}

@@

\section{Form Manipulation Language}
\label{sec: FML}
The discussion of the time discretisation so far has been general i.e.
not specific to any particular equation set. This is also true of the
implementation of the timestepping schemes. This enables flexibility
and reproducibility (**elaborate). This functionality is enabled
through the form manipulation language (FML) which is used to
associate a label to each term in the residual form defined in the
equation class. The timestepping methods are then able to reason about
which manipulations to apply to which forms by applying filters. Here
we will describe the objects required to define labels and filters
before illustrating this with a simple example(s?).

\subsection{Form Manipulation Language Objects}

The finite element discretisation of each term in an equation is
represented in Firedrake by a \texttt{ufl.Form} object. FML defines a
\texttt{Term} object that comprises the \texttt{ufl.Form} defining the
finite element discretisation of the term and a dictionary of
key-value pairs corresponding to the FML \texttt{Label} objects that
apply to this term. An FML \texttt{Label} object has a name and an
optional value. When called on a \texttt{ufl.Form} object, the
\texttt{Label} ** Labels can also be used to tag parts of the
form. For example, the \texttt{subject} label is used to tag the
\texttt{Function} in the \texttt{ufl.Form} that represents the
prognostic variable. These operations are composable, so more than one
label can be attached to a term.

The end result is a \texttt{LabelledForm} object that holds a list of
terms, pairing individual \texttt{ufl.Form} objects with an FML
\texttt{label}. Labelled forms can be added together, multiplied by
floats and have additional terms added. More complex manipulations are
enabled by the \texttt{label\_map} method that provides a way of
filtering the terms in the form based on their labels, applying
different mappings to terms that pass or fail the conditions defined
in the filter. A common operation might be to filter for terms that
have a certain label, such as \texttt{transport}, drop terms that do
not have this label and multiply those that do by the timestep. We
shall demonstrate this in the following example.

**describe methods has_label and get

\subsection{Example}

We demonstrate the use of FML by applying a split timestepping scheme
to the 1D advection diffusion equation
\begin{equation}\label{eq:1dadvdiff}
  q_t + cq_x - \kappa q_{xx} = 0,
\end{equation}
where $q$ is a scalar field, $c$ is the constant transporting velocity
and $\kappa$ is the constant diffusivity. The time discretisation
applies an explicit step of forward Euler to the transport term and
an implicit step of backward Euler to the diffusion term:
\begin{align}
  q^* &= q^n - c \Delta t q_x^n, \label{eq:explicit-advdiff} \\
  q^{n+1} - \kappa \Delta t q_{xx}^{n+1} &= q^*, \label{eq:implicit-advdiff}
\end{align}
where $q^n$ is the value of $q$ at time step $n$, $q^*$ is an
intermediate value for $q$ and $\Delta t$ is the timestep.

We illustrate the use of FML for this example in three steps: 1)
define and label the residual; 2) build the explicit and implicit
solvers; 3) run the timestepping loop. Since the intention of this
example is to demonstrate FML rather than the compatible finite
element discretisation, we use piecewise quadratic continuous Lagrange
polynomials, $P_2$, for the finite element discretisation as this
greatly simplifies the forms. The finite element discretisation of
equation \ref{eq:1dadvdiff} is then
\begin{equation}\label{eq:weak-advdiff}
  \int_\Omega vq \diff x + c \int_\Omega v q_x \diff x + \kappa \int_\Omega v_x q_x \diff x = 0 \quad \forall v \in P_2,
\end{equation}
where $v\in P_2$ is the test function and the diffusion term has been
integrated by parts.

%\begin{align}\label{eq:weak-advdiff}
%  \int_\Omega vq^* \diff x &= \int_\Omega vq^n \diff x - \Delta t c \int_\Omega v q_x^n \diff x, &\quad \forall v \in P^2, \\
%  \int_\Omega vq^{n+1} \diff x  + \Delta t \kappa \int_\Omega v_x q_x^{n+1} \diff x &= \int_\Omega vq^* \diff x, &\quad \forall v \in P^2,
%\end{align}
%where $v$ is the test function and the diffusion term has been
%integrated by parts.

Listing \ref{lst:define-residual} presents the code to define this
spatial discretisation of the residual with the $P_2$ function space
denoted by \texttt{V}. Here we omit the code that sets up the mesh,
function spaces and parameters $c$ and $\kappa$, although this can be
found in the supplementary materials. The first two lines define the
test function \texttt{v} and function \texttt{q}, corresponding to $v$
and $q$ in equation \ref{eq:weak-advdiff}. The next four lines define
the residual, i.e. the left hand side of equation
\ref{eq:weak-advdiff}, using the FML labels \texttt{time\_derivative},
\texttt{transport} and \texttt{diffusion} to tag the corresponding
terms. The final line labels the function \texttt{q} as the subject of
this residual. Note that the labels used for tagging just take the UFL
form as an argument and their value is simply the default, \texttt{True}, in
contrast to the \texttt{subject} label that has the value \texttt{q}.

\lstinputlisting[caption=blah, label={lst:define-residual}, language=python, firstline=24, lastline=31]{code/advection_diffusion_splitting_example.py}

To define the solvers for the explicit and implicit steps, we make use
of the predefined mappings \texttt{drop} and \texttt{keep}

\lstinputlisting[caption=blah, label={lst:define-explicit}, language=python, firstline=38, lastline=53]{code/advection_diffusion_splitting_example.py}

\lstinputlisting[caption=blah, label={lst:define-implicit}, language=python, firstline=56, lastline=70]{code/advection_diffusion_splitting_example.py}

\lstinputlisting[caption=blah, label={lst:timeloop}, language=python, firstline=56, lastline=70]{code/advection_diffusion_splitting_example.py}

@@

To include:
\begin{itemize}
\item terminator toy (Tim)
\item shallow water on different grids and with different HDiv
  families (Nell) e.g. Williamson 5 cubed sphere PV and Galewsky
  icosahedral sphere PV, high res (ref 6)
\item convergence with different RK schemes / multistage schemes using
  Williamson 2 or 6, L2 and Linf errors for D and u (2 plots, one L2,
  one Linf) (Alex)
\item vertical slice Euler - Straka, different orders, illustrates
  recovery for diffusion (Tom)
\item 3D moist bubble with rain (Tom/Alex)
\item spherical gravity wave, Boussinesq (Jemma)
\item spherical baroclinic wave, Euler - whatever people usually plot
  (horizontal slice pressure contours, temperature shading) (Dan)
\item Held Suarez??
\end{itemize}

@@

\subsection{Shallow water}
The standard suite of shallow water test cases originally proposed in
\citet{williamson1992standard} provides a common starting point for
testing new discretisations. This test suite is usually augmented by
the unstable barotropic jet described in
\citet{galewsky2004initial}. Here we present convergence results for a
variety of timestepping schemes applied to the steady state
geostrophic balance flow (test case 2 from
\citet{williamson1992standard}) and a demonstration of two different
grids and HDiv element families for shallow water flow over a mountain
(test case 5 from \citet{williamson1992standard}) and the unstable
barotropic jet \citep{galewsky2004initial}.

\subsubsection{Convergence results}

\subsubsection{Grids and HDiv elements}

\subsection{Compressible Boussinesq equations}

\subsection{Compressible Euler equations}

\subsubsection{Dense bubble in a vertical slice}

\subsubsection{3D moist rising bubble}

\subsubsection{Baroclinic wave on the sphere}

\subsubsection{Held Suarez}
